---
title: "HMB342 Assignment 3"
output: html_document
date: "2025-03-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
library(dplyr)#first read in unnecessary packages 
```

1. Read in the appropriate datasets and clean the data 

```{r}
pc_attachment <- read.csv("PC_Attachment.csv")
edv_low <- read.csv("EDV_Low_Urgency.csv")
edv_high <- read.csv("EDV_High_Urgency.csv")
```

Starting the cleaning process: 
```{r}
head(pc_attachment)
pc_attachment
names(pc_attachment)[1]<-paste("neighborhood_ID")
names(pc_attachment)[2]<-paste("neighborhood_name")
names(pc_attachment)[3]<-paste("total")
names(pc_attachment)[4]<-paste("attached")
names(pc_attachment)[5]<-paste("percent_attached")
names(pc_attachment)[6]<-paste("attached_female")
names(pc_attachment)[7]<-paste("attached_male")
names(pc_attachment)[8]<-paste("attached_under_19")
names(pc_attachment)[9]<-paste("attached_19_34")
names(pc_attachment)[10]<-paste("attached_35_49")
names(pc_attachment)[11]<-paste("attached_50_64")
names(pc_attachment)[12]<-paste("attached_65_79")
names(pc_attachment)[13]<-paste("attached_80_plus")
names(pc_attachment)[14]<-paste("uncertainly_attached")
names(pc_attachment)[15]<-paste("percent_uncertainly_attached")
names(pc_attachment)[16]<-paste("uncertainly_attached_female")
names(pc_attachment)[17]<-paste("uncertainly_attached_male")
names(pc_attachment)[18]<-paste("uncertainly_attached_19")
names(pc_attachment)[19]<-paste("uncertainly_attached_19_34")
names(pc_attachment)[20]<-paste("uncertainly_attached_35_49")
names(pc_attachment)[21]<-paste("uncertainly_attached_50_64")
names(pc_attachment)[22]<-paste("uncertainly_attached_65_79")
names(pc_attachment)[23]<-paste("uncertainly_attached_80_plus")
pc_attachment

pc_attachment <- pc_attachment %>%
  slice(-c(1,2)) %>%
  mutate(
    across(3:23, as.numeric), 
    across(1:2, as.character))
pc_attachment
```
```{r}
edv_low
edv_high
```


```{r}
edv_low
edv_low <- edv_low %>%
  slice(-c(1, 2, 3, 4, 5)) %>%
  mutate(
    across(3:132, as.numeric), 
    across(1:2, as.character))
edv_high <- edv_high %>%
  slice(-c(1, 2, 3, 4, 5)) %>%
  mutate(
    across(3:132, as.numeric), 
    across(1:2, as.character))

names(edv_high)[1]<-paste("neighborhood_ID")
names(edv_high)[2]<-paste("neighborhood_name")

names(edv_low)[1]<-paste("neighborhood_ID")
names(edv_low)[2]<-paste("neighborhood_name")

```
2. Now we want to look at how % of primary care attachment varies acrosss different ages and how the 
outcome of emergency department visits are altered (in both low & high urgency visits). 

A. Age Group 80+ 

```{r}
#first for the EDV Low Urgency Group 
names(edv_low)[123]<-paste("low_urgency_75_plus")
edv_low %>%
  select(neighborhood_name, low_urgency_75_plus) -> edv_low75

#now for the EDV High Urgency Group
names(edv_high)[123]<-paste("high_urgency_75_plus")
edv_high %>%
  select(neighborhood_name, high_urgency_75_plus) -> edv_high75

edv_high75


edv_low_high_75 <- full_join(edv_low75, edv_high75, 
            by = c("neighborhood_name"))

pc_attachment %>%
  select(neighborhood_name, total, attached_80_plus) %>%
  mutate(attached_percent_80_plus = ((attached_80_plus/total)*100)) -> pc_attachment_75
  


edv_attachment_75_comb <- full_join(edv_low_high_75, pc_attachment_75,
            by = c("neighborhood_name"))
  
```


```{r}
edv_attachment_75_comb %>%
  ggplot(aes(x = attached_percent_80_plus)) +
  geom_point(aes(y = high_urgency_75_plus, color = "high urgency")) +
    geom_point(aes(y = low_urgency_75_plus, color = "low urgency"))+
    scale_color_manual(name = "Emergency Department Visit Urgency", values = c("high urgency" = "red3", "low urgency" = "lightblue")) +
  theme_minimal() +
  labs(x = "Percent of Population Attached to Primary Care Physician", y = "Average Annual Rate of ED Visits (per 1000 population)")
```



B. Let's do this for the 65-79 age cohort!!



```{r}
#first for the EDV Low Urgency Group 
names(edv_low)[107]<-paste("low_urgency_65_79")
edv_low %>%
  select(neighborhood_name, low_urgency_65_79) -> edv_low_65_79

#now for the EDV High Urgency Group
names(edv_high)[107]<-paste("high_urgency_65_79")
edv_high %>%
  select(neighborhood_name, high_urgency_65_79) -> edv_high_65_79


edv_low_high_65_79 <- full_join(edv_low_65_79, edv_high_65_79, 
            by = c("neighborhood_name"))

pc_attachment %>%
  select(neighborhood_name, total, attached_65_79) %>%
  mutate(attached_percent_65_79 = ((attached_65_79/total)*100)) -> pc_attachment_65_79
  


edv_attachment_65_79_comb <- full_join(edv_low_high_65_79, pc_attachment_65_79,
            by = c("neighborhood_name"))
  
```

Visualizing the 65-79 Cohort:

```{r}
edv_attachment_65_79_comb %>%
  ggplot(aes(x = attached_percent_65_79)) +
  geom_point(aes(y = high_urgency_65_79, color = "high urgency")) +
    geom_point(aes(y = low_urgency_65_79, color = "low urgency"))+
    scale_color_manual(name = "Emergency Department Visit Urgency", values = c("high urgency" = "red3", "low urgency" = "lightblue")) +
  theme_minimal() +
  labs(x = "Percent of Population Attached to Primary Care Physician", y = "Average Annual Rate of ED Visits (per 1000 population)")
```

Looks like a strong relationship with attachment and the high urgency visit rate - is it significant? (pretty sure this 
is outide the scope of the assignment but im just curious) 


Fit Model to Data 

```{r}
lmmodel65_79 <- glm(high_urgency_65_79 ~ attached_percent_65_79, data = edv_attachment_65_79_comb, family = "gaussian")
summary(lmmodel65_79) #gaussian used bc data isn't count data but a rate 
AIC(lmmodel65_79) #goodness of model fit 

#looks like the relationship is significant (p-value) and the magnitude is moderately strong - for every 1% increase 
#in primary care attachment, the urgent emergency department visit rate for this age group decrease by 8.8
#lots of unexplained variability, but its ecological data so its giving ecological fallacy probs 

min_value <- min(edv_attachment_65_79_comb$attached_percent_65_79, na.rm = TRUE)
max_value <- max(edv_attachment_65_79_comb$attached_percent_65_79, na.rm = TRUE)

#creating the prediction values
new_data <- expand.grid(
  attached_percent_65_79 = seq(
    min(min_value),
    max(max_value),
    length.out = 100
  )
)

predictions <- predict(lmmodel65_79, newdata = new_data, type = "response")
new_data$predict <- predictions
  
```

#plotting predictive linear model over the data 
```{r}
ggplot(edv_attachment_65_79_comb, aes(x = attached_percent_65_79, y = high_urgency_65_79)) +
  geom_point(color = "darkblue", alpha = 0.6) +  # Scatter plot of observed data
  geom_line(data = new_data, aes(x = attached_percent_65_79, y = predict), color = "black", size = 1.2) +  # Prediction line
  labs(
    title = "Relationship between Primary Care Attachment and High Urgency ED Visits (65-79)",
    x = "Percent of Population Attached to Primary Care Physician",
    y = "Rate of High Urgency ED Visits per 1000 Population"
  ) +
  theme_minimal()
```
***Will talk about the results at the end of the document!!!


C. Let's do this for the 45-64 age cohort!!


```{r}
#first for the EDV Low Urgency Group 
names(edv_low)[90]<-paste("low_urgency_45_64")
edv_low %>%
  select(neighborhood_name, low_urgency_45_64) -> edv_low_45_64

#now for the EDV High Urgency Group
names(edv_high)[90]<-paste("high_urgency_45_64")
edv_high %>%
  select(neighborhood_name, high_urgency_45_64) -> edv_high_45_64


edv_low_high_45_64 <- full_join(edv_low_45_64, edv_high_45_64, 
            by = c("neighborhood_name"))

pc_attachment %>%
  select(neighborhood_name, total, attached_50_64) %>%
  mutate(attached_percent_45_64 = ((attached_50_64/total)*100)) -> pc_attachment_45_64
  


edv_attachment_45_64_comb <- full_join(edv_low_high_45_64, pc_attachment_45_64,
            by = c("neighborhood_name"))
  
```

Now Visualize the 45-64 Cohort: 

```{r}
edv_attachment_45_64_comb %>%
  ggplot(aes(x = attached_percent_45_64)) +
  geom_point(aes(y = high_urgency_45_64, color = "high urgency")) +
    geom_point(aes(y = low_urgency_45_64, color = "low urgency"))+
    scale_color_manual(name = "Emergency Department Visit Urgency", values = c("high urgency" = "red3", "low urgency" = "lightblue")) +
  theme_minimal() +
  labs(x = "Percent of Population Attached to Primary Care Physician", y = "AAverage Annual Rate of ED Visits (per 1000 population)")
```

D. Let's do this for the 35-49 Age Cohort!



```{r}
#first for the EDV Low Urgency Group 
names(edv_low)[73]<-paste("low_urgency_35_49")
edv_low %>%
  select(neighborhood_name, low_urgency_35_49) -> edv_low_35_49

#now for the EDV High Urgency Group
names(edv_high)[73]<-paste("high_urgency_35_49")
edv_high %>%
  select(neighborhood_name, high_urgency_35_49) -> edv_high_35_49


edv_low_high_35_49 <- full_join(edv_low_35_49, edv_high_35_49, 
            by = c("neighborhood_name"))

pc_attachment %>%
  select(neighborhood_name, total, attached_35_49) %>%
  mutate(attached_percent_35_49 = ((attached_35_49/total)*100)) -> pc_attachment_35_49
  


edv_attachment_35_49_comb <- full_join(edv_low_high_35_49, pc_attachment_35_49,
            by = c("neighborhood_name"))
  
```

Now Visualize the 45-64 Cohort: 

```{r}
edv_attachment_35_49_comb %>%
  ggplot(aes(x = attached_percent_35_49)) +
  geom_point(aes(y = high_urgency_35_49, color = "high urgency")) +
    geom_point(aes(y = low_urgency_35_49, color = "low urgency"))+
    scale_color_manual(name = "Emergency Department Visit Urgency", values = c("high urgency" = "red3", "low urgency" = "lightblue")) +
  theme_minimal() +
  labs(x = "Percent of Population Attached to Primary Care Physician", y = "AAverage Annual Rate of ED Visits (per 1000 population)")
```
E. Now the 20-34 Age Cohort: 


```{r}
#first for the EDV Low Urgency Group 
names(edv_low)[73]<-paste("low_urgency_20_34")
edv_low %>%
  select(neighborhood_name, low_urgency_20_34) -> edv_low_20_34

#now for the EDV High Urgency Group
names(edv_high)[73]<-paste("high_urgency_20_34")
edv_high %>%
  select(neighborhood_name, high_urgency_20_34) -> edv_high_20_34


edv_low_high_20_34 <- full_join(edv_low_20_34, edv_high_20_34, 
            by = c("neighborhood_name"))

pc_attachment %>%
  select(neighborhood_name, total, attached_19_34) %>%
  mutate(attached_percent_20_34 = ((attached_19_34/total)*100)) -> pc_attachment_20_34
  


edv_attachment_20_34_comb <- full_join(edv_low_high_20_34, pc_attachment_20_34,
            by = c("neighborhood_name"))
  
```

Now Visualize the 20-34 Cohort: 

```{r}
edv_attachment_20_34_comb %>%
  ggplot(aes(x = attached_percent_20_34)) +
  geom_point(aes(y = high_urgency_20_34, color = "high urgency")) +
    geom_point(aes(y = low_urgency_20_34, color = "low urgency"))+
    scale_color_manual(name = "Emergency Department Visit Urgency", values = c("high urgency" = "red3", "low urgency" = "lightblue")) +
  theme_minimal() +
  labs(x = "Percent of Population Attached to Primary Care Physician", y = "Average Annual Rate of ED Visits (per 1000 population)")
```
F. Now for the 0-19 Cohort: 

1) First Subset: 5-19

```{r}
#first for the EDV Low Urgency Group 
names(edv_low)[56]<-paste("low_urgency_5_19")
edv_low %>%
  select(neighborhood_name, low_urgency_5_19) -> edv_low_5_19

#now for the EDV High Urgency Group
names(edv_high)[73]<-paste("high_urgency_5_19")
edv_high %>%
  select(neighborhood_name, high_urgency_5_19) -> edv_high_5_19


edv_low_high_5_19 <- full_join(edv_low_5_19, edv_high_5_19, 
            by = c("neighborhood_name"))

pc_attachment %>%
  select(neighborhood_name, total, attached_under_19) %>%
  mutate(attached_percent_5_19 = ((attached_under_19/total)*100)) -> pc_attachment_5_19
  


edv_attachment_5_19_comb <- full_join(edv_low_high_5_19, pc_attachment_5_19,
            by = c("neighborhood_name"))
  
```

Now Visualize the 5-19 Cohort: 

```{r}
edv_attachment_5_19_comb %>%
  ggplot(aes(x = attached_percent_5_19)) +
  geom_point(aes(y = high_urgency_5_19, color = "high urgency")) +
    geom_point(aes(y = low_urgency_5_19, color = "low urgency"))+
    scale_color_manual(name = "Emergency Department Visit Urgency", values = c("high urgency" = "red3", "low urgency" = "lightblue")) +
  theme_minimal() +
  labs(x = "Percent of Population Attached to Primary Care Physician", y = "Average Annual Rate of ED Visits (per 1000 population)")
```
1) Second Subset: 0-4

```{r}
#first for the EDV Low Urgency Group 
names(edv_low)[39]<-paste("low_urgency_0_4")
edv_low %>%
  select(neighborhood_name, low_urgency_0_4) -> edv_low_0_4

#now for the EDV High Urgency Group
names(edv_high)[39]<-paste("high_urgency_0_4")
edv_high %>%
  select(neighborhood_name, high_urgency_0_4) -> edv_high_0_4


edv_low_high_0_4 <- full_join(edv_low_0_4, edv_high_0_4, 
            by = c("neighborhood_name"))

pc_attachment %>%
  select(neighborhood_name, total, attached_under_19) %>%
  mutate(attached_percent_0_4 = ((attached_under_19/total)*100)) -> pc_attachment_0_4
  


edv_attachment_0_4_comb <- full_join(edv_low_high_0_4, pc_attachment_0_4,
            by = c("neighborhood_name"))
  
```

Now Visualize the 0-4 Cohort: 

```{r}
edv_attachment_0_4_comb %>%
  ggplot(aes(x = attached_percent_0_4)) +
  geom_point(aes(y = high_urgency_0_4, color = "high urgency")) +
    geom_point(aes(y = low_urgency_0_4, color = "low urgency"))+
    scale_color_manual(name = "Emergency Department Visit Urgency", values = c("high urgency" = "red3", "low urgency" = "lightblue")) +
  theme_minimal() +
  labs(x = "Percent of Population Attached to Primary Care Physician", y = "Average Annual Rate of ED Visits (per 1000 population)")
```

3. Next, we want to put it all together: 


```{r}
edv_attachment_all_age_comb <- full_join(
  full_join(
    full_join(
      full_join(
        full_join(
          full_join(edv_attachment_0_4_comb, edv_attachment_5_19_comb, by = "neighborhood_name"),
          edv_attachment_20_34_comb, by = "neighborhood_name"),
        edv_attachment_35_49_comb, by = "neighborhood_name"),
      edv_attachment_45_64_comb, by = "neighborhood_name"),
    edv_attachment_65_79_comb, by = "neighborhood_name"),
  edv_attachment_75_comb, by = "neighborhood_name")

edv_attachment_all_age_comb<- edv_attachment_all_age_comb %>%
  rename(
    attached_percent_75_plus = attached_percent_80_plus
  )


```

```{r}
library(tidyr)
library(dplyr)
library(ggplot2)

# Assuming your merged dataframe is named edv_attachment_all_age_comb

# Pivot only the low and high urgency columns
edv_long <- edv_attachment_all_age_comb %>%
  pivot_longer(
    cols = c(
      "low_urgency_0_4", "high_urgency_0_4",
      "low_urgency_5_19", "high_urgency_5_19",
      "low_urgency_20_34", "high_urgency_20_34",
      "low_urgency_35_49", "high_urgency_35_49",
      "low_urgency_45_64", "high_urgency_45_64",
      "low_urgency_65_79", "high_urgency_65_79",
      "low_urgency_75_plus", "high_urgency_75_plus"
    ),
    names_to = c("urgency", "age_group"),
    names_pattern = "(low_urgency|high_urgency)_(.*)",
    values_to = "visits"
  )


```


```{r}
edv_long %>%
  ggplot(aes(x = age_group, y = visits, fill = urgency)) + 
  geom_boxplot(position = position_dodge(0.8)) + 
  labs(
    title = "Emergency Department Visits Across Age Groups", 
    x = "Age Group", 
    y = "Rate of Visits per 1000 Population", 
    fill = "Visit Type") + 
  theme_minimal()

```

What we see here is that there is a statistically significant difference in the rates of high urgency visits between the 0-4 and age groups over 65 and the other cohorts. There is also a statistically significant difference in the rate of low urgency visits for the 0-4 age group and all other age groups for low urgency visits. 


```{r}
# Pivot the attached percent 
edv_attachment_long <-edv_attachment_all_age_comb %>%
  pivot_longer(
    cols = c(
      "attached_percent_0_4",
      "attached_percent_5_19",
      "attached_percent_20_34",
      "attached_percent_35_49",
      "attached_percent_45_64",
      "attached_percent_65_79",
      "attached_percent_75_plus"
    ),
    names_to = "age_group",
    names_pattern = "attached_percent_(.*)",  # Extract age group part
    values_to = "attachment_percent"
  )

edv_combined <- edv_long %>%
  left_join(edv_attachment_long, by = "age_group")

```

```{r}
edv_combined %>%
  ggplot(aes(x = age_group, y = attachment_percent)) + 
  geom_boxplot(position = position_dodge(0.8), fill = "lightblue") + 
  labs(
    title = "Percent of Population with Primary Care Attachment", 
    x = "Age Group", 
    y = "Percent of Population Attached") + 
  theme_minimal()
```

What we see here is that the elderly populations are those with the lowest percentages on average with primary care, and the 65-79 and the 79+ age groups have statistically significant differences from the other age groups in terms of primary care attachment.


The results are significant for the 65-79 group. Eyeballing other groups it doesn't seem to be the case as much. What this could suggest is that this age group is more vulnerable to urgent health concerns than other groups likely for age related reasons, but that communities with higher percentages of primary care attachment have better health outcomes and fewer high urgency emergency department visits. It is possible that the reason we are not seeing this relationship as strongly with younger populations is that overall youth are healthier and have fewer urgent health issues than the elderly population, and having attachment to primary care physicians is not as crucial with younger ages. Our results suggest that elderly. With the exception of the 0-4 cohort, there seems to be almost a reverse relationship- maybe this can be attributed to po the fact that the attachment rates are not truly representative of this cohort since the attachment percentages were for ages 0-19, as opposed to 0-4 specifically, where it is likely that attachment rates differ and therefore the results are not representative of the true relationship. Similarly, since there is a bit of discrepancy between the attachment and ED visit datasets for the oldest age group (ie. attachment group's oldest age group is 80+, with the ED visits being 75+), this mismatch in ages likely confounded the results significantly. 
